Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-06-08 talk/icooolps2011/paper.tex
don't stress polymorphic inline caches anymore, focus on runtime feedback.

Meta-tracing JIT compilers can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to give the language implementor control over runtime
feedback. This restricted their performance. In this paper we describe the
mechanisms in PyPy's meta-tracing JIT that can be used to control
runtime feedback in language-specific ways. These mechanisms are flexible
enough to express classical VM techniques such as maps and polymorphic inline
caches.

@@

Together these hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such as maps and
polymorphic inline caches.

@@

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where knowing that a value is constant opens a lot of
optimization opportunities, even though it
could have different values in practice. In such a place, promotion can be used. The
typical reason to do that is if there is
a lot of computation depending on the value of one variable.

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives the user access to generalized inline caches. It requires of course compilation to JVM bytecode instead of simply writing an interpreter, predictability of performance across JVMs is also an open question.

We already explored promotion in other contexts, such as earlier versions of
PyPy's JIT.
%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

>>>>

Meta-tracing JIT compilers can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to give the language implementor control over runtime
feedback. This restricted their performance. In this paper we describe the
mechanisms in PyPy's meta-tracing JIT that can be used to control
runtime feedback in language-specific ways. These mechanisms are flexible
enough to express classical VM techniques such as maps and runtime type feedback.

@@

Together these hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such runtime type
feedback and maps.

@@

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where it would open a lot of optimization opportunities if a
variable were constant, even though it could have different values in
practice. In such a place, promotion can be used. The typical reason to do that
is if there is a lot of computation depending on the value of one variable.

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives
the user access to generalized inline caches. It requires of course compilation
to JVM bytecode instead of writing an interpreter.

We already explored promotion in other contexts, such as earlier versions of
PyPy's JIT.
%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
runtime type feedback (and also inline caching) techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be fed back into
compilation, not just types.
